Example Manuscript Template for a Data Analysis Project

### 2024-04-06 products/manuscript/Manuscript.qmd
Formatting and adding content.
Formatting and adding content.

---
title: "Example Manuscript Template for a Data Analysis Project"
subtitle: "Clarke Miller Project Proposal"
author: Clarke Miller
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.

You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(knitr))
```

# Summary/Abstract

# Introduction

## General Background Information

According to data provided by the World Health organization, seasonal respiratory illnesses are linked to over 650,000 deaths each year worldwide. In the United States the flu is responsible for approximately 42,000 deaths per year and ranks 13th most frequent cause of death.  As such, influenza is the subject of extensive public health surveillance and mediation efforts.

## Description of data and data source

Detailed data on covid and flu strains, outbreaks, and general epidemiological information is available via several sources. Data on these infectious diseases is available from sites such as GISAID (<https://gisaid.org/>) and the Global Health Data Exchange (<https://ghdx.healthdata.org/>). These databases are highly detailed and can provide data that includes information about lineage and geographic data. (See Figure 1 Below.)

```{r}

@@

## Questions/Hypotheses to be addressed

*State the research questions you plan to answer with this analysis.*

@@

I am certain that efforts like this already exist. My purpose for this project is to use it as a learning tool and to see if I can replicate those prior efforts.  Data to support this project is readily availble from multiple sources such as GISAID (<https://gisaid.org/>), the World Health Organization Global Influenza Programme (<https://www.who.int/tools/flunet>).

@@

# Methods

## Schematic of workflow

## Data aquisition

Data was acquired from the World Health Organization Global Influenza Programme. The data includes the excel files containing raw data (VIW_FNT_CM.xlsx) and a file containing the legend for the data (VIW_FLU_METADATA.csv).  This data set is one example of the data available for a project of this sort.  This specific data set includes information on 89 different parameters associated with influenza outbreaks as reported world-wide since about 1995.  The data includes time, region, type of flu, and many opther variables that are important for answereing the questions posed above.

## Data import and cleaning

```{r}
#Call a bunch of libraries.
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(broom)) 
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(readxl)) 
suppressPackageStartupMessages(library(dplyr)) 
suppressPackageStartupMessages(library(tidyr)) 
suppressPackageStartupMessages(library(skimr))
suppressPackageStartupMessages(library(gt))
suppressPackageStartupMessages(library(dslabs))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(kableExtra))

#Load excel data file
#Data file from WHO "VIW_FNT_CM.xlsx"

data_location <- here::here("data","raw-data","VIW_FNT_CM.xlsx")
WHO_data_flu <- readxl::read_excel(data_location)

#Save a rds version
save_data_location <- here::here("data","raw-data","VIW_FNT_CM.rds")
saveRDS(WHO_data_flu, file = save_data_location)

summary(WHO_data_flu)

```
Wow. That's a lot of data. Okay, let's filter the data a bit. I am still working out what my final analysis will look like and, frankly, doubting my coding abilities to produce a meaningful analysis. That being said, let's take a look.

For example, we might want to look at the total flu cases compared to the number of total cases of the A strains and the total number of the B strains in countries such as China, Australia, Canada, and France.  We also want to clean up the data by omitting any missing values. So we filter!

```{r}

WHO_data_flu <- WHO_data_flu %>% 
  select(WHOREGION, MMWR_WEEKSTARTDATE, COUNTRY_AREA_TERRITORY, INF_ALL, INF_A, INF_B)

summary(WHO_data_flu)

#| label: fig-schematic
#| fig-cap: "Figure 2: World Flu Burden by Year."
#| echo: FALSE
knitr::include_graphics(here("assets","plot1.png"))

#| label: fig-schematic
#| fig-cap: "Figure 3: World Flu Burden by Region."
#| echo: FALSE
knitr::include_graphics(here("assets","plot1B.png"))

@@

```{r}

#| label: fig-schematic
#| fig-cap: "Figure 4: World Flu Burden by Year for China Since Reporting Began."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2.png"))

@@

That's still a LOT of data. Notice the big gap around 2020-2022. That is due to the restrictions imposed by the COVID pandemic. Let's narrow this down to a more reasonable range... say to the beginning of the 2023-2024 flu season.

```{r}

#| label: fig-schematic
#| fig-cap: "Figure 5: World Flu Burden by Region."
#| echo: FALSE
knitr::include_graphics(here("assets","plot1C.png"))

```
These graphs help whittle down the data a bit.  For example, we can look at the stacked grpah in Figure 4 to see that over the course of the year the major flu variant shifted from influenza A to influenza B.  That being said, this format is not helpful for looking at the incidence over time per country or region.  To look at the sort of data we could do a stacked bar graph of the total flu counts per region/country or perhaps create a graph of the data using a simple scatter plot.

```{r}

#| label: fig-schematic
#| fig-cap: "Figure 6: World Flu Burden by Region Over the 2023-2024 Flu Season."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2B.png"))

```
Figure 6

## Statistical analysis
_Explain anything related to your statistical analyses._

@@

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

>>>>

---
title: "Geographic Spread of Seasonal Flu"
subtitle: "Clarke Miller MADA Project"
author: Clarke Miller
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}

# load a few R packages
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(knitr))

# Abstract:

# Introduction:

## General Background Information:

Influenza is an acute respiratory virus caused by one of several strains of virus.  Seasonal influenza is typically caused by circulating "A" and "B" strains of the disease.  Infection is accompanied by typical respiratory symptoms.  These may include cough, sore throat, and nasal discharge. Symptoms can also include abrupt onset of high-grade fever, muscle ache, loss of appetite, headache, and malaise.  After a typical course, influenza can go on to affect other organs such as the lungs, brain, and heart. Severe infections may require hospitalization.  Influenza infection can be much more severe for children, elderly adults, patients with chronic illnesses, and immunocompromised hosts.  According to data provided by the World Health organization, seasonal respiratory illnesses are linked to over 650,000 deaths each year worldwide. In the United States the flu is responsible for approximately 42,000 deaths per year and ranks 13th most frequent cause of death. As such, influenza is the subject of extensive public health surveillance and mediation efforts.**_[@Krammer-2018; @Moghadami-2017]_**

## Description of Data and Data Source:

Detailed data on covid and flu strains, outbreaks, and general epidemiological information is available via several sources. Data on these infectious diseases is available from sites such as GISAID (<https://gisaid.org/>) and the Global Health Data Exchange (<https://ghdx.healthdata.org/>). These databases are highly detailed and can provide data that includes information about lineage and geographic data. (See **Figure 1** Below.) Yearly surveillance data is also available from the World Health Organization Global Influenza Programme.

```{r, echo=FALSE, message=FALSE}

@@

```
**_Figure 1:_** *World Flu Burden by Clade.* Figure generated at GISAID.org.

## Questions/Hypotheses to be Addressed:

@@

I am certain that efforts like this already exist.  The best of these is software produced by Nextstrain (<https://nextstrain.org/>) which is capable of resolving complex time/geographic coupled with genetic data for reported viral infections.  Nextstrain offers real-time tracking of pathogen evolution for viruses such as SARS-CoV-2, seasonal flu, avian flu, mpox, RSV, and dengue.  In fact, the graphic used in Figure 1 (above) was made by GISAID using Nextstrain software.  My purpose for this project is to use this project as a learning tool and to see if I can replicate, in a small way, those prior efforts. Data to support this project is readily available from multiple sources such as GISAID (<https://gisaid.org/>), the World Health Organization Global Influenza Programme (<https://www.who.int/tools/flunet>).

@@

# Methods:

## Schematic of Workflow:

## Data Aquisition:

Data was acquired from the World Health Organization Global Influenza Programme. The data includes the excel files containing raw data (VIW_FNT_CM.xlsx) and a file containing the legend for the data (VIW_FLU_METADATA.csv). This data set is one example of the data available for a project of this sort. This specific data set includes information on 89 different parameters associated with influenza outbreaks as reported world-wide since about 1995. The data includes time, region, type of flu, and many other variables that are important for answering the questions posed above.

## Data import and cleaning:

Data was imported directly from the repository as an xlsx file. (VIW_FNT_CM.xlsx) The file was cleaned to remove missing data.  A .rds version of the data was saved (VIW_FNT_CM.rds).  Data that was filtered or summarized for graphs, tables, and for eventual statistical analysis were saved as individual .rds or csv files. All data analysis and statistical work was performed and documented in a separate .qmd file (workspace.qmd).

## Exploratory/Descriptive analysis:

In order to get a feel for the scale of data in the VIW_FNT_CM.xlsx one often needs to summarize or tabulate the data.  In this case, the data set is very large and summary tables exceedingly cumbersome. Given the size of the data set, it is best to begin with a graph of the incidence of influenza type A, type B, and the total number of reported cases for the duration of reported. (See **Figure 2.**)

```{r, echo=FALSE, message=FALSE}

#| label: fig-schematic
#| fig-cap: "Figure 2: World Flu Burden by Year."
#| echo = FALSE
knitr::include_graphics(here("assets","plot1.png"))

```
**_Figure 2:_** *World Flu Burden by Year.*

The graph reveals the extent of the data available.  The graph also reveals interesting features of the data set that may not have been readily apparent from if viewing the data in a tabular format.  Since 1995 the number of seasonal flu cases has followed a pattern that peaked during the winter months in the Northern hemisphere.  Notable features of this graph include the large jump in reported cases in the year 2010 and the large gap in reported cases in the year 2021.  The jump in cases on or about 2010 is due to increased participation in surveillance efforts and improved diagnostic techniques.  Hand in hand with these factors was an increasing awareness of the issue and consequent increased rate of reporting. **_[@Perdue-2012]_** The gap in the 2021 flu season was likely due to the measures taken to curb the SARS-CoV-2 pandemic through the 2020-2021 flu season.  Basic measures such as maintaining proper social distance, hand washing, public wearing of masks to prevent aerosol droplet spread, and other efforts were extremely effective at preventing the spread of seasonal flu. **_[@Rubin-2021]_**  Never the less, the amount of data reported worldwide since 1995 is massive.  In fact, the data is so dense that none of the categories of flu selected are visible as separate itmes in the bar graph.  Further filtering is obviously required.  For example, we might want to look at the total flu cases compared to the number of total cases of the A strains and the total number of the B strains in countries such as China, Australia, Canada, and France. It is also a good idea to clean up the data by omitting any missing values.  **Figure 3** depicts the data set filtered for information from China.

```{r, echo=FALSE, message=FALSE}

#| label: fig-schematic
#| fig-cap: "Figure 3: World Flu Burden by Year for China Since Reporting Began."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2.png"))

```
**_Figure 3:_** *World Flu Burden by Year for China Since Reporting Began.*

Restricting the data set to a single country doesn't seen to resolve the data in a useful manner.  Overall the data still shows the same basic structure and features as the entire data set.  It may be helpful to focus in on a single year.  **Figure 4** depicts the influenza surveillance data for the country of China for a single flu season.

```{r, echo=FALSE, message=FALSE}

```
**_Figure 4:_** *World Flu Burden by Year: China, 2023-2024 Flu Season.*

These graphs help whittle down the data a bit. For example, we can look at the stacked graph in **Figure 4** to see that over the course of the year the major flu variant shifted from influenza A to influenza B. That being said, this format is not helpful for looking at the incidence over time per country or per region and comparison of those regions.  To look at the data in that manner and, perhaps, look for time and geographical patterns a simple scatter plot may be helpful.  For example, a graph of total reported cases of flu for each WHO region over a single flu season may be revealing. (See **Figure 5**)

```{r, echo=FALSE, message=FALSE}

#| label: fig-schematic
#| fig-cap: "Figure 5: World Flu Burden by Region Over the 2023-2024 Flu Season."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2C.png"))

@@

**_Figure 5:_** *World Flu Burden by Region Over the 2023-2024 Flu Season.*

The resulting graph shows that the maxima for the three largest WHO flu regions, WPR, AMR, and EUR, do follow each other by approximately 3-4 weeks. The region encompassing Europe shows two distinct maxima. The first maxima is preceded by the maxima in the People's Republic of China WHO region by 3-4 weeks and corresponds to the largest maxima in the AMR data. The second is preceded by the maxima in the data from the WHO Americas Region by the same 3-4 week time frame. The data from the AMR region also show a second peak in the number of flu cases reported that follows the first EUR region peak by approximately 3-4 weeks.  Perhaps this indicates spread from EUR back to the AMR region?  As such there certainly seems to be a geographic/time component to the spread of the flu worldwide. (West to east in the northern hemisphere.) Further analysis of the available phylogeographic data may be able to pinpoint the origin of each specific strain.  In **Figure 5** it is also apparent that the relative magnitudes of the values of the WPR, AMR, and EUR regions overwhelm the data from the other WHO regions.  In this case it may be prudent to separate the graphs into separate graphs for WPR, AMR, EUR and AFR, EMR, SEAR. (See **Figure 6** and **Figure 7.**)

```{r, echo=FALSE, message=FALSE}

#| label: fig-schematic
#| fig-cap: "Figure 6: World Flu Burden by Region Over the 2023-2024 Flu Season."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2D.png"))

```
**_Figure 6:_** *World Flu Burden by Region Over the 2023-2024 Flu Season.*  WHO Regions: WPR, AMR, and EUR

```{r, echo=FALSE, message=FALSE}

#| label: fig-schematic
#| fig-cap: "Figure 7: World Flu Burden by Region Over the 2023-2024 Flu Season."
#| echo: FALSE
knitr::include_graphics(here("assets","plot2E.png"))

```
**_Figure 7:_** *World Flu Burden by Region Over the 2023-2024 Flu Season.* WHO Regions: AFR, EMR, SEAR

By splitting the graphs a great deal more detail can be discerned.  In addition to the putative west to east pattern that appeared for the WPR, AMR, and EUR WHO regions, there is a similar pattern for the SEAR and EMR regions.  An easy interpretation of this data would point to the seasonal flu beginning in the southeast Asia WHO region progressing to the eastern Mediterranean WHO region, then progressing to the WPR, AMR, and the EUR WHO regions respectively.  Of course, it is impossible to draw any conclusions from this data set and a single season.  To make a solid determination the data set would need to track not just time and geographic information, but connect the actual flu strains as they move around the globe.  This can be done with genetic data, i.e. DNA or RNA sequence data, from the flu cases reported.  In fact this is exactly the approach that the Nextstrain software takes.  In addition to presenting strong evidence of the central question of this research.  Further, graphing the data as opposed to attempting to genrate a summary table is a far more effective way to explore the data set.  As can be seen from **Table 1** (below), a tabular format doesn't highlight the patterns in the data in the same explicit manner as the graphic exploration.

```{r, echo=FALSE, message=FALSE}
data_location <- here::here("data","processed-data","flu_data_wide.rds")
flu_data_wide <- readRDS(data_location)

print(flu_data_wide)

@@

## Statistical Analysis:

*Explain anything related to your statistical analyses.*

@@

### 2024-04-06 products/manuscript/workspace.qmd
Formatting and adding content.
Formatting and adding content.

## Data import and cleaning
## Data import and cleaning

@@

```
Wow. That's a lot of data. Okay, let's filter the data a bit. I am still working out what my final analysis will look like and, frankly, doubting my coding abilities to produce a meaningful analysis. That being said, let's take a look.

@@

```
```{r}

@@

# Plot the data using a layered bar graph
plot3 <- ggplot(WHO_data_flu2, aes(x = MMWR_WEEKSTARTDATE)) +
  geom_bar(aes(y = INF_ALL, fill = "INF_ALL"), stat = "identity", color="black", position = "stack") + 
  geom_bar(aes(y = INF_A, fill = "INF_A"), stat = "identity", color="black", position = "stack") + 
  geom_bar(aes(y = INF_B, fill = "INF_B"), stat = "identity", color="black", position = "stack") + 
  labs(title = "Influenza Cases Reported to WHO",
       subtitle = "Country = China, 2022-2024 Seasons",
       x = "Week",
       y = "Number of Positive Specimens",
       fill = "Influenza Type") + 
  theme_minimal()

@@

# Plot the region axis data using a layered bar graph
plot1C <- ggplot(WHO_data_fluB, aes(x = WHOREGION)) +
  geom_bar(aes(y = INF_ALL, fill = "INF_ALL"), stat = "identity", color="black", position = "stack") + 
  labs(title = "Influenza Cases Reported to WHO",
       subtitle = "Country = all, 2022-2024 Season",
       x = "WHO Region",
       y = "Number of Positive Specimens",
       fill = "Influenza Type") + 
  theme_minimal()

@@

>>>>

## Data import and cleaning

## Data import and cleaning

@@

Wow. That's a lot of data. Okay, let's filter the data a bit. I am still working out what my final analysis will look like and, frankly, doubting my coding abilities to produce a meaningful analysis. That being said, let's take a look.

@@

@@

# Plot the data using a layered bar graph
plot3 <- ggplot(WHO_data_flu2, aes(x = MMWR_WEEKSTARTDATE)) +
  geom_bar(aes(y = INF_ALL, fill = "INF_ALL"), stat = "identity", color="black", position = "stack") + 
  geom_bar(aes(y = INF_A, fill = "INF_A"), stat = "identity", color="black", position = "stack") + 
  geom_bar(aes(y = INF_B, fill = "INF_B"), stat = "identity", color="black", position = "stack") + 
  labs(title = "Influenza Cases Reported to WHO",
       subtitle = "Country = China, 2023-2024 Seasons",
       x = "Week",
       y = "Number of Positive Specimens",
       fill = "Influenza Type") + 
  theme_minimal()

@@

# Plot the region axis data using a layered bar graph
plot1C <- ggplot(WHO_data_fluB, aes(x = WHOREGION)) +
  geom_bar(aes(y = INF_ALL, fill = "INF_ALL"), stat = "identity", color="black", position = "stack") + 
  labs(title = "Influenza Cases Reported to WHO",
       subtitle = "Country = all, 2023-2024 Season",
       x = "WHO Region",
       y = "Number of Positive Specimens",
       fill = "Influenza Type") + 
  theme_minimal()

@@

library(lubridate)
library(dplyr)

summarized_data_work <- summarized_data 
summarized_data_work <- summarized_data_work %>% mutate(MMWR = format(MMWR_WEEKSTARTDATE , "%Y-%m"))

print(summarized_data_work)

monthly_summary <- summarized_data_work %>% group_by(WHOREGION, MMWR) %>%
summarize(TOT_FLU = sum(TOT_INF_ALL))

print(monthly_summary)

monthly_summary1 <- monthly_summary %>% filter(MMWR > '2023-07')
monthly_summary1 <- monthly_summary1 %>% filter(MMWR > '2023-07')

selected_regions3 <- c("WPR", "AMR", "EUR", "AFR", "SEAR", "EMR")
monthly_summary2 <- subset(monthly_summary1, WHOREGION %in% selected_regions3)

monthly_summary2 <- monthly_summary2 %>%
  group_by(WHOREGION, MMWR)

flu_data_wide <- monthly_summary2 %>% pivot_wider(names_from = MMWR, values_from = TOT_FLU)

print(flu_data_wide)

#Save a rds version
save_data_location <- here::here("data","processed-data","flu_data_wide.rds")
saveRDS(flu_data_wide, file = save_data_location)

data_location <- here::here("data","processed-data","flu_data_wide.rds")
flu_data_wide <- readRDS(data_location)

flu_data_table <- flu_data_wide |> 
  gt() |> 
  cols_label(
    WHOREGION = 'WHO REGION',
    `2023-08` = 'AUG',
    `2023-09` = 'SEPT',
    `2023-10` = 'OCT',
    `2023-11` = 'NOV',
    `2023-12` = 'DEC',
    `2024-01` = 'JAN',
    `2024-02` = 'FEB',
  ) |> 
  tab_spanner(
    label = md('**Total Flu Cases Per Month for Each WHO Region.**'),
    columns = c(`2023-08`,`2023-09`,`2023-10`,`2023-11`, `2023-12`,`2024-01`,`2024-02`)
  )

flu_data_table

flu_data_table1 <- flu_data_wide |> 
  gt() |> 
  cols_label(
    `2023-08` = "AUG",
    `2023-09` = "SEPT",
    `2023-10` = "OCT",
    `2023-11` = "NOV",
    `2023-12` = "DEC",
    `2024-01` = "JAN",
    `2024-02` = "FEB"
  ) |> 
  tab_spanner(
    label = md('**Total Flu Cases Per Month for Each WHO Region.**'),
    columns = c(`2023-08`, `2023-09`, `2023-10`, `2023-11`, `2023-12`, `2024-01`, `2024-02`)
  )

flu_data_table1
